---
title: "p8105_hw3_ml5018"
author: "Luan Mengxiao"
date: 2023-10-05
output: github_document
---

This is a R Markdown document for homework 3.

Load the package to be used for problem solving.

```{r setup, message = FALSE}
library(tidyverse)

options(tibble.print_min = 5)

knitr::opts_chunk$set(
  fig.width = 6,
  fig.asp = .6,
  out.width = "90%"
)

theme_set(theme_minimal() + theme(legend.position = "right"))

options(
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill = "viridis"
)

scale_colour_discrete = scale_colour_viridis_d
scale_fill_discrete = scale_fill_viridis_d
```

# Problem 1

## load the data

This problem uses the Instacart data. Load the data from the `p8105.datasets` using the following code chunk.

```{r loading_1}
library(p8105.datasets)
data("instacart")
```

## describe the data

The goal is to do some exploration of this dataset. To that end, write a short description of the dataset, noting the size and structure of the data, describing some key variables, and giving illustrative examples of observations.

While the original data is quite extensive, the dataset loaded from `p8105.datasets` represents a cleaned and limited version. The dataset contains 1,384,617 observations of 131,209 unique users, where each row in the dataset is a product from an order. There is a single order per user in this dataset.

The dataset consists of `r nrow(instacart)` observations of `r ncol(instacart)` variables, which form a total of `r nrow(instacart) * ncol(instacart)` data points. The dataset is organized in a longer form that makes it tidy and easy for computation using codes.

Some of the key vaiables of the dataset include `r names(instacart)`.

The 15 variables contain information about the order, product and customer, as listed in details below:

* order_id: order identifier
* product_id: product identifier
* add_to_cart_order: order in which each product was added to cart
* reordered: 1 if this prodcut has been ordered by this user in the past, 0 otherwise
* user_id: customer identifier
* eval_set: which evaluation set this order belongs in (Note that the data for use in this class is exclusively from the “train” eval_set)
* order_number: the order sequence number for this user (1=first, n=nth)
* order_dow: the day of the week on which the order was placed
* order_hour_of_day: the hour of the day on which the order was placed
* days_since_prior_order: days since the last order, capped at 30, NA if order_number=1
* product_name: name of the product
* aisle_id: aisle identifier
* department_id: department identifier
* aisle: the name of the aisle
* department: the name of the department

And some of the key characteristics of these variables are summarized below.

```{r description_1}
skimr::skim(instacart)
```

The code chunk below gives out a few illustrative examples of observations.

```{r illustration_1}
print(instacart, width = Inf)
```

Then, do or answer the following (commenting on the results of each):

How many aisles are there, and which aisles are the most items ordered from?

```{r aisle}
instacart |>
  group_by(aisle_id, aisle) |>
  summarize(n_obs = n()) |>
  arrange(desc(n_obs))
```

There are `r instacart |> pull(aisle_id) |> max()` aisles in total, among which the most items come from `fresh vegetables` and `fresh fruits`, with little difference between the amounts of these two aisles.

Make a plot that shows the number of items ordered in each aisle, limiting this to aisles with more than 10000 items ordered. Arrange aisles sensibly, and organize your plot so others can read it.

```{r aisle_plot}
instacart |>
  group_by(aisle) |>
  summarize(n_obs = n()) |>
  filter(n_obs > 10000) |>
  mutate(aisle = fct_reorder(aisle, n_obs)) |>
  ggplot(aes(x = aisle, y = n_obs)) +
  geom_col(fill = "blue") +
  coord_flip() +
  labs(
    title = "Number of Items Ordered at Each Aisle",
    x = "Aisle name",
    y = "Number of items ordered",
    caption = "Note: limited to aisles with more than 10000 items ordered"
  )
```

A column plot is used to visualize the data we get. To show aisle names clearly without overlapping, the x and y axis need to be flipped.

It can be seen from the plot that "fresh vegetables" and "fresh fruits" possess the most itmes ordered, which verifies the result we get in the lase part.

Make a table showing the three most popular items in each of the aisles “baking ingredients”, “dog food care”, and “packaged vegetables fruits”. Include the number of times each item is ordered in your table.

```{r table_aisle}
get_table_aisle = function(aisle_name){
  result = 
    instacart |>
    filter(aisle == aisle_name) |>
    group_by(product_name, product_id) |>
    summarize(n_obs = n()) |>
    arrange(desc(n_obs)) |>
    mutate(aisle = aisle_name) |>
    select(aisle, everything())
  return(result[1:3, ])
}

table_popular = 
  bind_rows(
    get_table_aisle("baking ingredients"), 
    get_table_aisle("dog food care"), 
    get_table_aisle("packaged vegetables fruits")
    )
table_popular |>
  knitr::kable()
```

The protocol for each of the three aisles are the same, thus to facilitate the code, define a function `get_table_aisle` to implement the main process and then bind the resulting rows from the function.

Separated tables can also be generated using this function.

```{r}
get_table_aisle("baking ingredients") |>
  select(-aisle) |>
  knitr::kable()
get_table_aisle("dog food care") |>
  select(-aisle) |>
  knitr::kable()
get_table_aisle("packaged vegetables fruits") |>
  select(-aisle) |>
  knitr::kable()
```

It can be seen from the tables that there are siginificant differences in the number of most ordered items from different aisles.

Make a table showing the mean hour of the day at which Pink Lady Apples and Coffee Ice Cream are ordered on each day of the week; format this table for human readers (i.e. produce a 2 x 7 table).

```{r table_hour}
get_table_hour = function(order_name){
  result = 
    instacart |>
    filter(product_name == order_name) |>
    group_by(order_dow) |>
    summarize(
      mean_hour_of_the_day = mean(order_hour_of_day)
    ) |>
    mutate(product_name = order_name,
           order_dow = case_match(
             order_dow,
             0 ~ "Sunday",
             1 ~ "Monday",
             2 ~ "Tuesday",
             3 ~ "Wednesday",
             4 ~ "Thursday",
             5 ~ "Friday",
             6 ~ "Saturday"
           )) |>
    pivot_wider(
      names_from = order_dow,
      values_from = mean_hour_of_the_day
    ) |>
    select(product_name, everything())
  return(result)
}

table_hour = bind_rows(
  get_table_hour("Pink Lady Apples"),
  get_table_hour("Coffee Ice Cream")
)

table_hour |> knitr::kable()
```

The method is similar to the one used in last problem, defining a function `get_table_hour` to carry out the main process and then applying the function to get a combined table.

The required 2 x 7 table can also be obtained using this function.

```{r}
get_table_hour("Pink Lady Apples") |>
  select(-product_name) |>
  knitr::kable()
get_table_hour("Coffee Ice Cream") |>
  select(-product_name) |>
  knitr::kable()
```

It can be concluded from the tables that there is no siginificant differnce in the ordered number of the two products, all of which lie between 10 and 16.

# Problem 2

## load the data

This problem uses the BRFSS data. Load the data from the `p8105.datasets` package.

```{r loading_2}
data("brfss_smart2010")
```

## clean the data

First, do some data cleaning:

* format the data to use appropriate variable names;
* focus on the “Overall Health” topic
* include only responses from “Excellent” to “Poor”
* organize responses as a factor taking levels ordered from “Poor” to “Excellent”

```{r cleaning}
brfss_smart2010 = 
  brfss_smart2010 |>
  janitor::clean_names() |>
  filter(topic == "Overall Health",
         response %in% c("Excellent", "Very good", "Good", "Fair", "Poor")) |>
  mutate(response = factor(response, ordered =TRUE, 
                           levels = c("Poor", "Fair", "Good", "Very good", "Excellent")))
```

## Explore the data

Using this dataset, do or answer the following (commenting on the results of each):

In 2002, which states were observed at 7 or more locations? What about in 2010?

```{r state}
get_state = function(year_number){
  result = 
    brfss_smart2010 |>
    filter(year == year_number) |>
    group_by(locationabbr) |>
    summarize(n_obs = n_distinct(locationdesc)) |>
    filter(n_obs >= 7)
  return(result)
}

brfss_2002 = get_state(2002)
brfss_2010 = get_state(2010)
```

In 2002, `r pull(brfss_2002, var = 1)` were observed at 7 or more locations.

In 2010, `r pull(brfss_2010, var = 1)` were observed at 7 or more locations.

The `n_distinct` function is used to count the number of observations in each location, and a function `get_state` is defined to get result from the same process more simlpy.

Construct a dataset that is limited to `Excellent` responses, and contains, year, state, and a variable that averages the `data_value` across locations within a state. Make a “spaghetti” plot of this average value over time within a state (that is, make a plot showing a line for each state across years – the `geom_line` geometry and `group` aesthetic will help).

```{r excellent}
brfss_excellent = 
  brfss_smart2010 |>
  filter(response == "Excellent") |>
  group_by(locationabbr, year) |>
  summarize(mean_data_value = mean(data_value, na.rm = TRUE)) |>
  select(year, locationabbr, mean_data_value)

brfss_excellent |>
  ggplot(aes(x = year, y = mean_data_value, group = locationabbr)) +
  geom_line() +
  labs(
    title = "Spaghetti Plot for `mean_data_value` over time within a state",
    x = "year",
    y = "mean_data_value",
    caption = "Data from BRFSS dataset"
  )
```

There are missing values in data_value column, thus `na.rm` is applied to omit these NAs when calculating mean.

From the spaghetti plot we can see the change of mean data value within states over time, which is quite complicated to read from. It can be adjusted by using different colors for different groups, as shown below.

```{r excellent_color}
brfss_excellent |>
  ggplot(aes(x = year, y = mean_data_value, color = locationabbr)) +
  geom_line() +
  theme(legend.position = "right")
  labs(
    title = "Spaghetti Plot for `mean_data_value` over time within a state",
    x = "year",
    y = "mean_data_value",
    caption = "Data from BRFSS dataset"
  )
```

Make a two-panel plot showing, for the years 2006, and 2010, distribution of `data_value` for responses (“Poor” to “Excellent”) among locations in NY State.

```{r two_panel}
brfss_smart2010 |>
  filter(year %in% c(2006,2010),
         locationabbr == "NY") |>
  ggplot(aes(x = response, y = data_value, fill = locationdesc)) +
  geom_col() +
  facet_grid(.~year) +
  theme(axis.text.x = element_text()) +
  labs(
    title = "`data_value` to `response` Column Plots for 2006 and 2010",
    x = "response",
    y = "data_value",
    caption = "data from BRFSS dataset"
  )

brfss_smart2010 |>
  filter(year %in% c(2006,2010),
         locationabbr == "NY") |>
  ggplot(aes(x = data_value, fill = response)) +
  geom_density(alpha = 0.5) +
  facet_grid(.~year) +
  theme(axis.text.x = element_text()) +
  labs(
    title = "`data_value` to `response` Density Plots for 2006 and 2010",
    x = "response",
    y = "density",
    caption = "data from BRFSS dataset"
  )

brfss_smart2010 |>
  filter(year %in% c(2006,2010),
         locationabbr == "NY") |>
  ggplot(aes(x = response, y = data_value)) +
  geom_boxplot(aes(fill = response)) +
  facet_grid(.~year) +
    theme(axis.text.x = element_text()) +
  labs(
    title = "`data_value` to `response` Box Plots for 2006 and 2010",
    x = "response",
    y = "data_value",
    caption = "data from BRFSS dataset"
  )
```

Some of the settings were set up in the first code chunk, applying for all the plots generated. A column, density and box plot are drawn to compare between different years and response groups.

It can be concluded from the plots that the overall patterns of data value distribution for response are quite similar in the two years chosen, despite the different degree of disparation. And for different resonse groups, the "poor" group possesses a significantly different density and distribution compared with other groups, which tend to overlap with each other on the density plot.

# Problem 3

## loading the data

Load, tidy, merge, and otherwise organize the data sets. Your final dataset should include all originally observed variables; exclude participants less than 21 years of age, and those with missing demographic data; and encode data with reasonable variable classes (i.e. not numeric, and using factors with the ordering of tables and plots in mind).

```{r import}
nhanes_accel = 
  read_csv("./data/nhanes_accel.csv") |>
  janitor::clean_names() |>
  pivot_longer(
    min1:min1440,
    names_to = "min",
    names_prefix = "min",
    values_to = "value"
  )
nhanes_covar = read_csv("./data/nhanes_covar.csv", skip = 4) |>
  janitor::clean_names()

nhanes_df = 
  full_join(nhanes_covar, nhanes_accel) |>
  drop_na(sex, age, bmi, education) |>
  filter(age >= 21) |>
  mutate(sex = case_match(sex, 1 ~ "male", 2 ~ "female"),
         sex = factor(sex, levels = c("male", "female")),
         education = case_match(education,
                                1 ~ "Less than high school",
                                2 ~ "High school equivalent",
                                3 ~ "More than high school"),
         education = factor(education, 
                            levels = c("Less than high school", "High school equivalent", "More than high school")))
```

## explore the data

Produce a reader-friendly table for the number of men and women in each education category, and create a visualization of the age distributions for men and women in each education category. Comment on these items.

```{r table_education}
nhanes_df |>
  group_by(sex, education) |>
  summarize(n_obs = n()) |>
  pivot_wider(
    names_from = education,
    values_from = n_obs
  ) |>
  knitr::kable()
```

There is no significant difference between number of men and women in eduacation group "Less than high school" and "more than high school", but the number of men in "High school equivalent" group is significantly larger than that of women.

The table can also be flipped if different strategies are used for `pivot_wider` function, using names from `sex` instead of `education`.

```{r table_sex}
nhanes_df |>
  group_by(sex, education) |>
  summarize(n_obs = n()) |>
  pivot_wider(
    names_from = sex,
    values_from = n_obs
  ) |>
  knitr::kable()
```

Traditional analyses of accelerometer data focus on the total activity over the day. Using your tidied dataset, aggregate across minutes to create a total activity variable for each participant. Plot these total activities (y-axis) against age (x-axis); your plot should compare men to women and have separate panels for each education level. Include a trend line or a smooth to illustrate differences. Comment on your plot.

```{r comparison}
nhanes_df |>
  group_by(seqn) |>
  summarize(sum = sum(value)) |>
  left_join(nhanes_df) |>
  select(sex, age, education, sum) |>
  ggplot(aes(x = age, y = sum, color = sex)) +
  geom_point() +
  geom_smooth(alpha = .5) +
  facet_grid(education~.) +
  labs(
    title = "Total Activities against Age for Males and Females",
    x = "age",
    y = "Total activity",
    caption = "Total activity is gained by aggregation across minutes"
  )

nhanes_df |>
  group_by(seqn) |>
  summarize(sum = sum(value)) |>
  left_join(nhanes_df) |>
  select(sex, age, education, sum) |>
  ggplot(aes(x = age, y = sum, color = sex)) +
  geom_line(alpha = .5) +
  geom_smooth(alpha = .5) +
  facet_grid(education~.) +
  labs(
    title = "Total Activities against Age for Males and Females",
    x = "age",
    y = "Total activity",
    caption = "Total activity is gained by aggregation across minutes"
  )
```

A point plot is drawn due to the disparation of the data, from which we can see that the despite the fluctuation, total activity in "Less than high school" group tends to decrease as age increases, while the other two groups show more stablility across ages.

Accelerometer data allows the inspection activity over the course of the day. Make a three-panel plot that shows the 24-hour activity time courses for each education level and use color to indicate sex. Describe in words any patterns or conclusions you can make based on this graph; including smooth trends may help identify differences.

```{r three_panel}
nhanes_df |>
  ggplot(aes(x = min, y = value, color = sex)) +
  geom_point(alpha = .5) +
  geom_smooth() +
  facet_grid(education~.) +
  labs(
    title = "24-Hour Activity Time Courses",
    x = "time",
    y = "activity",
    caption = "Data from nhanes dataset"
  ) 

nhanes_df |>
  ggplot(aes(x = min, y = value, color = sex)) +
  geom_line(alpha = .5) +
  geom_smooth() +
  facet_grid(education~.) +
  labs(
    title = "24-Hour Activity Time Courses",
    x = "time",
    y = "activity",
    caption = "Data from nhanes dataset"
  ) 
```

A line plot is applied to manifest the activity changing pattern according to time. The alpha value is set because of the overlapping between different sex groups in one plot.

It can be seen that while the pattern of male in all three education groups look similar, the activity of female in "More than high school" group shows siginificant difference compared with the other two groups.